A Simple Type Theory for Category Theory

### 2022-04-27 paper/paper.tex
some more intro

@@

@@

More experienced practitioners of category theory in programming
language theory find that for modeling computational effects such as
general recursion or indexed-state, or modeling logics for
parametricity, gradual typing, or step-indexed logical relations
involves not just ordinary category theory, but more general notions
such as enriched, internal or multi-categories. However, model
construction theorems and especially soundness and completeness
theorems in these areas will require their own, even more complex
versions of the same common lemmas of category theory, in addition to
the domain-specific results such as existence of fixed points.

At this level of generality, the proofs can be difficult to
understand, as they often involve some combination of the original
category theoretic proof with some additional side-conditions that
everythinng is in fact appropriately step-indexed, continuous etc.
%
One approach to taming this complexity is to prove these general
theorems at a higher level of generality, for instance, proving them
in some abstract 2-category of enriched/internal categories, functors
and transformations.
%
For proofs in generalized category theory specifically, the notion of
a \emph{virtual equipment} has been shown to be occupy an algebraic
sweet-spot of being complex enough to express most basic constructions
(functors, profunctors, natural transformations, adjunctions, kan
extensions, etc) but also general enough to encompass internal,
enriched and ``generalized multi-categories'', which include ordinary
flavors of multi-categories that model intuitionistic and
substructural type theories, but also a great deal more, including
double categories, metric spaces and topological spaces.
%
However, there are two difficulties with this approach.
%
First, the algebraic arguments can be difficult to understand compared
to the original argument for ordinary category theory, which is an
obstacle to wider adoption of this powerful approach.
%
Second, the algebraic structure of a virtual equipment is quite
unwieldy to manipulate. The abstraction of a generalized natural
transformation is a kind of ``multi-2-cell'' with many inputs and a
single output. In diagrammatic form these are rendered as

@@

\subsection{Category Theory with Profunctors}

The syntax of Judgment Theory gives not just a syntax for the familiar
trio of category, functor and natural transformation, but additionally
a concept that is not usually highlighted in introductory category
theory texts: the profunctor. A simple intuition is that profunctors
are to functors as relations are to functions. Indeed, the syntax for
judgment theory has a straightforward semantics in the category of
sets and relations.

The reason profunctors are not commonly introduced is that they are
seen as an auxiliary notion: a profunctor $R : C \pto D$ is
typically defined as a functor $C^{\textrm{op}}\times D \to
\textrm{Set}$. However this definition requires two concepts that are
somewhat problematic: the opposite category $C^\textrm{op}$ and the
category of (small) sets $\textrm{Set}$. However, these notions do not
necessarily exist in all of our intended applications. First, we are
interested in semantics where are objects are \emph{multicategories}
where morphisms have multiple inputs but only a single output. In this
case the opposite of a multicategory is not a multicategory at all but
instead some kind of ``co-multicategory'' with multiple outputs but
one input. Second, having a universe object like $\textrm{Set}$ is also

\subsection{Applications to Enriched, Internal and Multicategory Theory}

The syntax of judgment theory is an internal language for
\emph{virtual equipments}. These are a kind of joint generalization of
double categories and multicategories.

Notably, the definition of a category in \cref{sec:internal} can be
applied as an extremely general notion of category, including internal
categories and enriched categories. This general notion is often
called a ``generalized multi-category'' and encompasses an astonishing
range of mathematical structures, including categories, topological
spaces, metric spaces, etc.

\subsection{Applications to Generalized Type Theory}

While we mostly focus on the semantics in category theory, due to the
tight connection between category theory and the semantics of type
theory, there are direct applications to the study of ``abstract type
theory'' as well. We demonstrate how we can model single-variable
simple type theories with products and coproducts within SCT. This is
very much in the spirit of logical frameworks \cite{TODO}, since we
can faithfully model the terms of an object theory inside of our
metatheory. Since we have models of our metatheory in virtual double
categories, we establish sound semantics as well. Put succinctly,
category theory itself is the logical framework for the study of type
theories.

@@

To demonstrate the utility of SCT, we proceed as follows
\begin{enumerate}
\item In \cref{sec:syntax} we introduce the syntax of Simple Category
  Theory.
\item In \cref{sec:synthetic} we show how common theorems of category
  theory such as the (co)-Yoneda lemma, equivalence of different
  formulations of adjunctions, and preservation of limits can all be
  cleanly reproduced in SCT.
\item In \cref{sec:internal}, we show how we can use SCT to define
  \emph{internal categories}, which allows us to prove additional
  theorems that are not formalizable as synthetic theorems.
\item In \cref{sec:semantics}, we show that the syntax of SCT is sound
  for models in an arbitrary virtual double category with
  corresponding universal constructions.
\item In \cref{sec:discussion}, we discuss related type theories and
  potential extensions to model multi-variable and higher-order
  categorical constructions.
\end{enumerate}

@@

Next, we develop the basic model theory for SCT. Because SCT includes
not just a syntax for category theory, but is also indexed by an
ordinary context of dependent type theory $\Gamma$, a model of SCT is
not given by a \emph{single} virtual double category, but instead by
an \emph{indexed family} of VDCs, parameterized by an object of a
category that interprets dependent type theory. We call such a
structure a \emph{VDC-hyperdoctrine}. We then show how to construct a
from any VDC, a VDC-hyperdoctrine indexed by the category of sets,
which has the intuitive reading that any of the VDC judgments $\Gamma
\vdash J$ is interpreted as simply a family of $J$-data parameterized
by elements of the set $\Gamma$.

@@

@@

String Diagrams?

@@

An orthogonal extension to these would be to consider adding a
\emph{discretization} modality $|\cat C|$ that gives the discrete
category on the underlying set of objects of $\cat C$. This would give
us the ability to formalize concepts like ``function on objects'' or
``unnatural transformation'' without the need to define internal
categories. This could be modeled by adding a modal distinction
$\alpha:^{\#}\cat C$ for ``crisp'' dependence on an object of $\cat
C$, i.e., dependence merely on the object of $\cat C$ with no
functoriality/naturality conditions. This would then be internalized
by the discretization modality.

\paragraph{As a tool for Formal Type Theory}

While we can model ``unary'' type theories using profunctors, to model
theories like simple intuitionistic, linear or bunched type theories
we need a way to talk about ``profunctors of arbitrarily many
arguments''. This can be accomplished using \emph{multicategory}
theory, which models the structure of the context with its structural
rules as a monad of ``contexts and substitutions''. A suitable notion
of monad on a virtual double category has been developed and was in
fact the original motivation for the concept of a virtual double
category. However while an abstract monad of contexts $T$ is enough to
define what a $T$-multicategory internal to a virtual equipment is, it
does not appear possible to define the multiplicative connectives as a
universal property, i.e., to define when the multicategory is
\emph{representable}.

A recent development in the formalization of substructural type
theories is to use monoidal connectives at the meta-level when
defining multiplicative connectives\cite{woodsatkeyesop2022}. This
suggest one approach to extending the theory to support the
formalization of multiplicative universal properties would be, in
adherence to the microcosm principle, to add linear judgments and
multiplicative connectives to the metalevel.

>>>>

This work is merely a first step in this direction, as there are still
three major components of category theory: opposite categories,
functor categories and ``universes'' such as the category of sets or
categories that are not nicely axiomatizable in the current system.
%
Nontheless, a great deal of category theory can be carried out in this
more limited setting, and we can still take advantage of duality at a
meta level.

@@

First, we review the 4 basic notions of our theory of categories in
the traditional formulation. A category $\cat C$ is a set of objects
$\cat C_0$ with a span of arrows $\cat C_0 \leftarrow \cat C_1
\rightarrow \cat C_0$ that pick out the source and target of each
arrow, such that there are composition and identity arrows with
appropriate laws. A functor $F : \cat C \to \cat D$ consists of a
function on objects $F_0 : \cat C_0 \to \cat D_0$ and a function on
morphisms $F_1 : \cat C_1 \to \cat D_1$ that is compatible with source
and target and preserves identity and composition.

A profunctor $R$ from $\cat C$ to $\cat D$ is a span $\cat C_0
\leftarrow R \rightarrow \cat D_0$ that is compatible with
\emph{left}-composition by $\cat C$ and \emph{right}-composition by
$\cat D$. If we think of categories as generalizations of (ordered)
sets then functors generalize (monotone) functions and profunctors
generalize (monotone) \emph{relations}. We think of an element $r \in
R$ with $s(r) = c$ and $t(r) = d$ as a ``proof'' that $c R d$ or some
kind of ``heteromorphism'' from $c$ to $d$. The compatibility with
composition says that these relations are down-closed in $C$ and
up-closed in $D$, or that heteromorphisms can be composed with
homomorphisms.
%

@@

We can abstract the algebraic structure of categories, functors,
profunctors, and natural transformations as a \emph{double category}.
%
A double category is a generalization of categories where we have a
collection of objects $C_0$, and \emph{two separate} notions of
morphism, which we call vertical $C_v$ and horizontal $C_h$ over the
same objects $C_0$ which each have their own identity and composition,
and finally a notion of 2-cell that is most easily described

@@

%% \subsection{Category Theory with Profunctors}

%% The syntax of Judgment Theory gives not just a syntax for the familiar
%% trio of category, functor and natural transformation, but additionally
%% a concept that is not usually highlighted in introductory category
%% theory texts: the profunctor. A simple intuition is that profunctors
%% are to functors as relations are to functions. Indeed, the syntax for
%% judgment theory has a straightforward semantics in the category of
%% sets and relations.

%% The reason profunctors are not commonly introduced is that they are
%% seen as an auxiliary notion: a profunctor $R : C \pto D$ is
%% typically defined as a functor $C^{\textrm{op}}\times D \to
%% \textrm{Set}$. However this definition requires two concepts that are
%% somewhat problematic: the opposite category $C^\textrm{op}$ and the
%% category of (small) sets $\textrm{Set}$. However, these notions do not
%% necessarily exist in all of our intended applications. First, we are
%% interested in semantics where are objects are \emph{multicategories}
%% where morphisms have multiple inputs but only a single output. In this
%% case the opposite of a multicategory is not a multicategory at all but
%% instead some kind of ``co-multicategory'' with multiple outputs but
%% one input. Second, having a universe object like $\textrm{Set}$ is also

%% \subsection{Applications to Enriched, Internal and Multicategory Theory}

%% The syntax of judgment theory is an internal language for
%% \emph{virtual equipments}. These are a kind of joint generalization of
%% double categories and multicategories.

%% Notably, the definition of a category in \cref{sec:internal} can be
%% applied as an extremely general notion of category, including internal
%% categories and enriched categories. This general notion is often
%% called a ``generalized multi-category'' and encompasses an astonishing
%% range of mathematical structures, including categories, topological
%% spaces, metric spaces, etc.

%% \subsection{Applications to Generalized Type Theory}

%% While we mostly focus on the semantics in category theory, due to the
%% tight connection between category theory and the semantics of type
%% theory, there are direct applications to the study of ``abstract type
%% theory'' as well. We demonstrate how we can model single-variable
%% simple type theories with products and coproducts within SCT. This is
%% very much in the spirit of logical frameworks \cite{TODO}, since we
%% can faithfully model the terms of an object theory inside of our
%% metatheory. Since we have models of our metatheory in virtual double
%% categories, we establish sound semantics as well. Put succinctly,
%% category theory itself is the logical framework for the study of type
%% theories.

@@

\begin{enumerate}
\item In \cref{sec:syntax} we introduce the syntax of Simple Category
  Theory.

\item In \cref{sec:examples} we show how to use our syntax for
  synthetic category theory. Much of this is just rendering well-known
  2-categorical/double categorical constructions in our type theoretic
  syntax to demonstrate the ``naive''/synthetic character of proofs
  and constructions in SCT. We additionally give some examples of how
  this can be applied to model a recent formulation of modal dependent
  type theory.

%% \item In \cref{sec:synthetic} we show how common theorems of category
%%   theory such as the (co)-Yoneda lemma, equivalence of different
%%   formulations of adjunctions, and preservation of limits can all be
%%   cleanly reproduced in SCT.
%% \item In \cref{sec:internal}, we show how we can use SCT to define
%%   \emph{internal categories}, which allows us to prove additional
%%   theorems that are not formalizable as synthetic theorems.

\item In \cref{sec:semantics}, we show that the syntax of SCT is sound
  for models in an arbitrary virtual double category with
  corresponding universal constructions.

\item In \cref{sec:discussion}, we discuss related type theories and
  potential extensions to model multi-variable and higher-order
  categorical constructions.
\end{enumerate}

We have aimed that the syntax and examples sections should be
accessible to anyone familiar with basic category theory and
willingness to understand profunctors and (co-)ends.
%
The semantics section is by necessity more advanced and may be skipped
or skimmed by readers who are willing to trust the mathematical
foundations of the system.

@@

Next, we develop the basics of the model theory for SCT. Because SCT
includes not just a syntax for category theory, but is also indexed by
an ordinary context of dependent type theory $\Gamma$, a model of SCT
is not given by a \emph{single} virtual double category, but instead
by an \emph{indexed family} of VDCs, parameterized by an object of a
category that interprets dependent type theory. We call such a
structure a \emph{VDC-hyperdoctrine}. We then show how to construct a
from any VDC, a VDC-hyperdoctrine indexed by the category of sets,
which has the intuitive reading that any of the VDC judgments $\Gamma
\vdash J$ is interpreted as simply a family of $J$-data parameterized
by elements of the set $\Gamma$.

@@

\paragraph{Calculi for Presheaves/Profunctors}

David Jaz Myers String Diagrams

Coend calculus

@@

@@

An orthogonal extension to these would be to consider adding a
\emph{discretization} modality $\Delta\cat C$ that gives the discrete
category on the underlying set of objects of $\cat C$. This would give
us the ability to formalize concepts like ``function on objects'' or
``unnatural transformation'' without the need to define internal
categories. This could be modeled by adding a modal distinction
$\alpha:^{\#}\cat C$ for ``crisp'' dependence on an object of $\cat
C$, i.e., dependence merely on the object of $\cat C$ with no
functoriality/naturality conditions. This would then be internalized
by the discretization modality.

A third approach would be to extend the calculus to model general
monoidal closed structures and more generally multicategories, which
provides semantics of modal substructural type theories.
%
This can be modeled by extending the ``universe'' $\Set$ to be a
monoidal closed category, which induces a monoidal closed structure on
profunctors by the Day convolution product \cite{...}.
%
Similar developments along these lines have already been done for
coend calculus \citep{coend-calculus} and in formalization of
substructural type theories \citep{atkey-woods}.

%% A recent development in the formalization of substructural type
%% theories is to use monoidal connectives at the meta-level when
%% defining multiplicative connectives\cite{woodsatkeyesop2022}. This
%% suggest one approach to extending the theory to support the
%% formalization of multiplicative universal properties would be, in
%% adherence to the microcosm principle, to add linear judgments and
%% multiplicative connectives to the metalevel.

%% \paragraph{As a tool for Formal Type Theory}

%% While we can model ``unary'' type theories using profunctors, to model
%% theories like simple intuitionistic, linear or bunched type theories
%% we need a way to talk about ``profunctors of many arguments''. This
%% can be accomplished using \emph{multicategory} theory, which models
%% the structure of the context with its structural rules as a monad of
%% ``contexts and substitutions''. A suitable notion of monad on a
%% virtual double category has been developed and was in fact the
%% original motivation for the concept of a virtual double
%% category. However while an abstract monad of contexts $T$ is enough to
%% define what a $T$-multicategory internal to a virtual equipment is, it
%% does not appear possible to define the multiplicative connectives as a
%% universal property, i.e., to define when the multicategory is
%% \emph{representable}.

\paragraph{Semantics}

Our definition of a double category with displays is suggestive of a
simple generalization. A double category is a monoid object in the
double category of $Span(Cat)$, and a double category with displays
furthermore uses certain diagrams involving profunctors, which also
form a double category $Prof$. This suggests that the algebraic
structure of a double category with displays is naturally modeled in
some kind of \emph{triple category}, modeling the three types of
morphisms: functors, profunctors and spans of functors.
